SparseIDS: Learning Packet Sampling with Reinforcement Learning

### 2020-02-04 paper_rl/paper.tex
Cleanup

The Critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$. For this it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$. It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following value loss function:
\begin{align}
\begin{split}
l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{classification},t}\right)^2 \\
+ & \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\right)^2
\end{split}
\end{align}

Now that the reward metrics and the Critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each timestep $t$ as follows:
\begin{align}
\begin{split}
U_t =& \left(R_{\text{classification},t} + \alpha \cdot R_{\text{sparsity},t}\right) \\
- & \left(v_{\text{classification},t} + \alpha \cdot v_{\text{sparsity},t}\right)
\end{split}
\end{align}

This means that we want the reward to be higher than the Critic's expectation. Note that it would be also possible to unify both reward metrics in one and have the Critic have only one output, but we consider it to be more interpretable to have two separate outputs, which can be inspected by an operator.

The Actor outputs a probability distribution at timestep $t$, which is sampled to determine the number of packets that should be skipped. It aims to change the distribution so that actions which result in a higher reward are chosen more frequently, maximizing the utility function. However, this strategy could lead to the Actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward. This would be equivalent to the Actor being stuck in a local maximum. Thus, the Actor not only tries to change the distribution so that the actions become more optimal (higher reward) but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.

Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled. $a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen, $a_t = 3$ means that two are skipped and so forth. Besides wanting to make actions which result in a high reward more likely, the actor also seeks to optimize the entropy $H(\cdot)$ of the probability distribution so that it keeps exploring and doesn't get stuck with a suboptimal policy.

@@

\begin{figure}
\centering
  \includegraphics[width=0.8\columnwidth]{img/rnn-sampling.pdf}
  \caption{SparseIDS framework. Three deep networks are shown: an LSTM classifier, An Actor network and a Critic network. The Actor takes actions based on the feedback from the the Critic network and the classifier. The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.}
  \label{fig:sampling}
\end{figure}

@@

Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during testing but instead take the mean of the distribution for the continuous case and the mode for the discrete one. This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore since they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.

>>>>

The critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$. For this it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$. It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following value loss function:
\begin{align}
\begin{split}
l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{classification},t}\right)^2 \\
+ & \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\right)^2
\end{split}
\end{align}

Now that the reward metrics and the critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each timestep $t$ as follows:
\begin{align}
\begin{split}
U_t =& \left(R_{\text{classification},t} + \alpha \cdot R_{\text{sparsity},t}\right) \\
- & \left(v_{\text{classification},t} + \alpha \cdot v_{\text{sparsity},t}\right)
\end{split}
\end{align}

This means that we want the reward to be higher than the critic's expectation. Note that it would be also possible to unify both reward metrics in one and have the critic have only one output, but we consider it to be more interpretable to have two separate outputs, which can be inspected by an operator.

The actor outputs a probability distribution at timestep $t$, which is sampled to determine the number of packets that should be skipped. It aims to change the distribution so that actions which result in a higher reward are chosen more frequently, maximizing the utility function. However, this strategy could lead to the actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward. This would be equivalent to the actor being stuck in a local maximum. Thus, the actor not only tries to change the distribution so that the actions become more optimal (higher reward) but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.

Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the actor $\theta_a$ the actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled. $a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen, $a_t = 3$ means that two are skipped and so forth. Besides wanting to make actions which result in a high reward more likely, the actor also seeks to optimize the entropy $H(\cdot)$ of the probability distribution so that it keeps exploring and doesn't get stuck with a suboptimal policy.

@@

\begin{figure}
\centering
  \includegraphics[width=0.8\columnwidth]{img/rnn-sampling.pdf}
  \caption{SparseIDS framework. Three deep networks are shown: an LSTM classifier, An actor network and a critic network. The actor takes actions based on the feedback from the the critic network and the classifier. The critic network tracks the actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the actor network.}
  \label{fig:sampling}
\end{figure}

@@

Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during testing but instead take the mean of the distribution for the continuous case and the mode for the discrete one. This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore since they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore. Furthermore, the critic is only necessary during training for the reward of the actor. During testing, it can be disabled, saving computational power.
